---
title: "Week 10: Exploring DHS data with ggplot2"
output:
  prettydoc::html_pretty:
    theme: architect
    toc: true
author: "YOUR NAME HERE"
date: "2023-12-05"
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Load necessary packages
```{r}
pacman::p_load(
  janitor,    # for data analysis utilities
  here,       # to set the working directory
  haven,      # for reading Stata files
  tidyverse   # for data manipulation and visualization
)
```

*NOTE: Please read through and run the code in the demo Rmd BEFORE starting this exercise.*

# Data Importation and cleaning

## Step 1: Choose and import select columns

The code chunk below imports the same variables as presented in the demo. However, we want to augment this by adding five variables related to malaria-knowledge.

Scan through the data dictionary file and determine which five variables correspond to the following agreement statements:

-   Malaria can be fully cured by medicine
-   Malaria can lead to death
-   No worry about malaria due to easy treatment
-   Know people sick with malaria
-   Only weak children can die from malaria

Then, incorporate these variables into the data import code below. 

HINT - these variables can be found on pages 225-226 of the data dictionary PDF.

**Note**: make sure ALL letters in the variable names are in lower case. If the data dictionary says "S1108BA" you should change the 3 letters to lowercase like this: "s1108ba" when you enter it in `col_select`.

```{r}
# Import the .dta file and add the 5 malaria knowledge variables
ir_ng_18_raw <- 
  haven::read_dta(here("data/NGIR7BFL_mini.DTA"),
                  col_select = c(
                    # Basic information
                    v006, v007, v009, v010, v012, v013, 
                    v024, v025, v106, v130, v190,                     
                    
                    # Literacy and media
                    v155, v156, v157, v158, v159, 
                    
                    # Fertility and reproductive timing
                    v201, v212, v213, v511, v531, 
                    
                    "WRITE YOUR CODE HERE" 
                    
                    ))
```


## Step 2: Assign variable names

Assign descriptive names to the newly imported variables and then execute the code below.

```{r}
# Rename malaria knowledge variables
ir_ng_18_renamed <- 
  ir_ng_18_raw %>% 
  rename(age = v012,
         age_group = v013,
         region = v024,
         urban_rural = v025,
         highest_educ = v106,
         religion = v130,
         literacy = v155,
         wealth_index = v190,
         num_kids = v201,
         age_first_birth = v212,
         age_first_cohabitation = v511,
         age_first_sex = v531, 
         "WRITE YOUR CODE HERE" 
         )
```

## Step 3: Reveal coded values

With the suitable function from the `haven` package, transform all variables into regular R factors.

```{r}
# Add function to convert from labelled data to regular factors
ir_ng_18_clean <- 
  ir_ng_18_renamed %>% 
  "WRITE YOUR CODE HERE"
```

# Comparing distributions

## Histograms, boxplots, and bar plots for continuous distributions

a.  Choose one of the continuous variables from the dataset: `age`, `num_kids`, `age_first_sex`, `age_first_cohabitation` or `age_first_birth.` Then, generate a simple histogram of this variable. Make sure to provide the necessary labels.

```{r}
# Simple histogram
"WRITE YOUR CODE HERE"
```

b.  Now, incorporate the `fill` aesthetic to color the bars of your histogram by levels of a categorical variable. Map one of the new malaria knowledge variables to fill - `aes(..., fill = MALARIA_VAR)`, and use `position` adjustments to create a side-by-side (dodged) histogram. This graph should show how the responses to this question differ at the various levels of the chosen continuous variable.

```{r}
# Histogram with position argument
"WRITE YOUR CODE HERE"
```

c. The method above isn't ideal. Visualize the same comparison with a boxplot instead of a histogram. Adjust the colors, labels, titles, themes, etc. as you please.

```{r}
# Boxplot comparing groups
"WRITE YOUR CODE HERE"
```

## Barplots for discrete distributions 

a.  Select one of the categorical variables from the dataset. Proceed to generate a simple bar plot to display the distribution of this variable. 

```{r}
# Simple bar plot
"WRITE YOUR CODE HERE"
```


b.  Next, add the `fill` aesthetic, `aes(fill = MALARIA_VAR)`, with `position` adjustments to generate a PERCENT-STACKED bar plot. This graph should illustrate how the answers to the malaria knowledge question change at the different levels of the chosen categorical variable.

```{r}
# Percent stacked bar plot
"WRITE YOUR CODE HERE"
```

c. Choose two OTHER categorical variables and created a GROUPED bar chart. Adjust the colors, labels, titles, themes, etc. as you please.

```{r}
# Grouped bar chart
"WRITE YOUR CODE HERE"
```


# Bonus challenge

Pick one of the plots from above to customize further, then implement the below customizations.

## Facets 

Another way of comparing distributions by levels of a categorical variable is **faceting**. For example:

```{r} 
ggplot(data = mtcars, aes(x = mpg, y = disp)) +
  geom_point() +
  facet_wrap(~cyl)
```

Implement something similar in your plots, faceting by a selected categorical variable.


## Custom color scales

Rather than just let {ggplot2} pick your fill colors for you, you can implement a custom palette using `scale_color_manual()` (or `scale_fill_manual()`) and the `values` argument. 

To generate a custom palette, go to https://mycolor.space and enter this hex code, which is the GRAPH Network's log color: `#6DA5AF`. Then generate a suitable palette based on this color. 

Copy the hex codes and use them in the `scale_color_manual()` function as done in the example below:

```{r}
# Scatterplot with custom colors
ggplot(data = mtcars, aes(x = mpg, y = disp, color = factor(cyl))) +
  geom_point() +
  scale_color_manual(values = c("#4C5B5C", "#6B8E7A", "#EFF2A9")) +
  facet_wrap(~cyl) +
  theme_light()
```

